---
title: "PGA Project"
author: "Tyler Gorecki"
date: "7/5/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(highlight=FALSE)
knitr::opts_chunk$set(fig.width = 4.75, fig.height = 3, fig.align = 'center')
```

```{r, echo=FALSE, include=FALSE}

library(dplyr)
library(ggplot2)
library(BSDA)

pga_data <- read.csv("/Users/tylergorecki/Desktop/Past UVA Semesters/2022  Spring/STAT 3080/Project/pgastats20-21.csv")
clubheadspeed <- read.csv("/Users/tylergorecki/Desktop/Past UVA Semesters/2022  Spring/STAT 3080/Project/clubheadspeed21.csv")
ballspeed <- read.csv("/Users/tylergorecki/Desktop/Past UVA Semesters/2022  Spring/STAT 3080/Project/ballspeed21.csv")
strokesgained <- read.csv("/Users/tylergorecki/Desktop/Past UVA Semesters/2022  Spring/STAT 3080/Project/StrokesGained.csv")

clubheadspeed <- clubheadspeed %>% 
  rename(NAME = PLAYER.NAME)

ballspeed <- ballspeed %>% 
  rename(NAME = PLAYER.NAME)

strokesgained <- strokesgained %>% 
  rename(NAME = PLAYER.NAME)

big_pga <- merge(pga_data, 
                 merge(clubheadspeed, 
                       merge(ballspeed, strokesgained, 
                             by = "NAME"), 
                       by = "NAME"), 
                 by = "NAME")

big_pga <- big_pga %>% 
  rename(AVG.CHS = AVG..x, FASTEST.CLUBHEAD.SPEED = FASTEST.SPEED.x, 
         AVG.BS = AVG..y, FASTEST.BALL.SPEED = FASTEST.SPEED.y, 
         SLOWEST.CLUBHEAD.SPEED = SLOWEST.SPEED.x, SLOWEST.BALL.SPEED = SLOWEST.SPEED.y)

big_pga <- big_pga %>% 
  mutate(AVERAGE.SG.T = TOTAL.SG.T/RNDS, AVERAGE.SG.T2G = TOTAL.SG.T2G/RNDS, 
         AVERAGE.SG.P = TOTAL.SG.P/RNDS)

big_pga <- merge(big_pga, strokesgained, by = "NAME")

big_pga <- big_pga %>%
  mutate(AnyWin = if_else(WINS == 0, "NoWins", "Win(s)"))
```

# What is the most important component of success on the PGA Tour? 

## Background and Data Description

Golf is an incredibly difficult sport both mentally and physically. With the 
analytics boom impacting all sports, golf has followed suit with new approaches
to golf ball/ club technology and in playing strategy. I collected the PGA Tour 
data by transferring the statistics from both the ESPN (1) 
and PGA Tour (2) websites into an excel spreadsheet. This data was then converted into 
csv files and merged in R by the name of each golfer in the dataset. The ESPN data 
included traditional statistics for the golfers - statistics like final ranking, number 
of rounds played, number of wins, average driving distance, and 13 others. The 
PGA Tour data contained more advanced and modern analytics such as club head speed, 
ball speed, and strokes gained. Strokes gained is a modern way of comparing 
golfers directly with each other (3). For every round in each tournament, the 
net number of strokes gained by all players that day will be zero. If a golfer 
played better than the rest of the field, their strokes gained will be positive 
and if they didn't play well compared to the field, it will be negative. My goal 
is to determine what the best approach is to perform well and win on the PGA Tour. 

## More Specifically...

Does driving distance or fairway accuracy matter more? Is greens in regulation 
or putting the most important? Some players main focus is to make sure they always 
try to hit the fairway to give them a better lie to hit the green on their following 
shots. Others attempt to hit the ball as far as they can to give themselves a shorter 
distance to the green. I want to research which strategy is better correlated with 
success among PGA Tour golfers on the PGA Tour. I also want to analyze how much 
impact greens in regulation (the percentage of shots hitting the green in no more 
than two shots under par - ex: by the second shot on a par four) and putting have 
in the difference between the top golfers and the rest. 

### Potential Issues

When comparing the data between both sites, I realized that there was a slight 
difference in the number of rounds for each golfer in their 2020-21 seasons. The 
number of rounds was slightly higher from the PGA Tour website (2) and I believe this 
is due to the fact that they included their postseason tournament in their final 
statistics, while ESPN (1) did not. The numbers aren't significantly different, but 
because of this, I will not be comparing any 'total' statistics. Regardless, there 
is enough data for each golfer that the 'average' statistics can be trusted for 
appropriate analysis for this dataset. 

## Scoring

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=4.5, fig.align='center'}
big_pga %>%
  ggplot(aes(x = SCORE, fill = AnyWin)) +
  geom_bar(stat = 'count', position = 'dodge') +
  geom_text(stat = 'count', aes(label = ..count..), 
            position = position_dodge(.1), vjust = -0.5, size = 2) +
  labs(title = "Typical scores for golfers with vs without win(s)", 
       x = "Average score per round", 
       y = "Number of golfers") +
  theme(plot.title = element_text(hjust = 0.5))
```

This graph shows what the typical scores are of golfers that 
won a tournament this season and that of the golfers who didn't. It 
is expected, but from this numerical summary the golfers who had at least one win on the 
PGA Tour during the 2020-21 season tended to average a lower average score over every 
round they played during the season. From this visualization, I know that 2/3, or 
28/42 of the winners in this dataset averaged a score of 70.5 or less per round 
over the season. Looking further at this average score, about 36% of the golfers 
that averaged this low of a score won at least one tournament, while only about 
12% who averaged above this threshold won a tournament. This shows that winning 
on tour typically isn't a fluke and the most consistent golfers win tournaments 
more times than not. This is a good baseline when first looking at this data to 
understand that there is a consistent trend that can be seen among PGA Tour winners. 

## Driving

```{r, echo=FALSE, warning=FALSE, message=FALSE}
big_pga %>%
  ggplot(aes(x = DACC, y = AVERAGE.x)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = "Strokes gained vs. driving accuracy", 
       x = "Percentage of fairways hit off the tee", 
       y = "Average number of strokes gained per round") +
  theme(plot.title = element_text(hjust = 0.5))
```

While there is a slight positive correlation in this relationship, this plot 
shows that driving accuracy isn't a true indicator of success on the PGA Tour 
unless the percentage of fairways hit is above 70% or below 55%, which is rare. 
For the data points within the 55-70% range, the average number of strokes gained 
per round can almost be considered random between -1 and 1.5 strokes gained. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
big_pga %>%
  ggplot(aes(x = DDIS, y = AVERAGE.x)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = "Strokes gained vs. driving distance", 
       x = "Distance of golf ball hit off the tee", 
       y = "Average number of strokes gained per round") +
  theme(plot.title = element_text(hjust = 0.5))
```

When looking at the relationship between strokes gained and driving distance, 
while there is a slightly higher positive correlation between the variables, 
a similar sense of randomness can be seen regardless of the average driving 
distance. From these plots, I feel comfortable saying that as long as the golfer 
drives the ball in play, they should just focus on giving themselves the best 
opportunity to hit the green on their next shot. For some, hitting out of the 
fairway is a big priority, but others believe that regardless of lie, it is always 
better to be as close to the green as possible for hitting an approach shot. 

## Greens in regulation and putting

```{r, echo=FALSE, warning=FALSE, message=FALSE}
big_pga %>%
  ggplot(aes(x = GIR, y = AVERAGE.x)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = "Strokes gained vs. greens in regulation", 
       x = "Percentage of greens hit in regulation", 
       y = "Average number of strokes gained per round") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
big_pga %>%
  ggplot(aes(x = PUTTS, y = AVERAGE.x)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title = "Strokes gained vs. putting", 
       x = "Average number of putts per hole", 
       y = "Average number of strokes gained per round") +
  xlim(1.68, 1.83) +
  theme(plot.title = element_text(hjust = 0.5))
```

These two plots both have seemingly strong correlations showing relationships 
between strokes gained and greens in regulation percentage and strokes gained and 
average number of putts per hole. 

For players with at least 67% of greens hit in regulation, there are still 
multiple who have an average number of strokes gained per round below -0.5, 
meaning they tend to score lower than the rest of the players on tour. For 
putting though, of the golfers averaging less than 1.75 putts per hole, there is 
only one that has an average strokes gained significantly less than 0. While 
we're talking about an average of .08 average putts per hole and about 1.44 per 
round, this could be the difference in a golfer winning or losing a tournament. 
I'm not saying that golfers aren't trying to set themselves up in the best 
situations for putting, but if there's a most important aspect to a golfer's 
game I could identify, it would be approach shots. Better approach shots leads 
to less putts and less putts leads to more success on the PGA Tour. 

```{r, include=FALSE}
## Statistical Test
### One-Sample t-test
#I will be testing the if mean of the average strokes gained putting of the top 
#20 golfers is equal to or greater than the mean of the average strokes gained 
#of all golfers in the 2020-21 PGA Tour season. As stated earlier, the net strokes 
#gained is always zero, so that will be the value of the mean for my null hypothesis. 
#Because my dataset does not contain every single golfer that participated in the 
#season of interest, I do not know the population standard deviation and thus will 
#perform a one sample t-test. For the assumptions of this test, the data is continuous, there 
#are no significant outliers in the sample, and the data is approximately normally distributed 
#(shown in qqplot below). The strokes gained statistic is somewhat dependent one other golfers
#(must sum to zero), but I feel comfortable continuing with this 
#test and trusting the results because the physical putting of each golfer is independent 
#and that's what causes the differences within this statistic. 

qqnorm(big_pga$AVERAGE.SG.P, pch = 1, frame = FALSE)
qqline(big_pga$AVERAGE.SG.P, col = "steelblue", lwd = 2)

### Hypotheses
#H0: mu = 0 vs. HA: mu > 0

### Test code


topPutting <- big_pga %>% filter(RK <= 20) %>% select(AVERAGE.SG.P)
t.test(topPutting, alternative = 'greater', mu = 0, conf.level = 0.95)


### Result and Conclusion
#Test stat: t = 3.8056; p-value: p = 0.0005972

#Because the p-value is less than 0.05, there is enough evidence to reject the 
#null hypothesis and suggest that the true 
#mean strokes gained of top 20 golfers on the PGA Tour in the 2020-21 season is 
#greater than zero. This means that the top golfers are better putters on average 
#than all PGA Tour golfers as a whole. The results from the test 
#even suggests that the true mean for the top 20 golfers is actually 0.237. This 
#means that the top golfers gain about .237 strokes per round on the rest of the 
#field from putting alone (almost a stroke per tournament, which are usually four 
#rounds). Putting is clearly important for success on the PGA Tour. 

## Further application of the data
#This test can be used to explain the putting of top golfers during the 2020-21 PGA Tour 
#season, but I believe it can be generalized past just the population of golfers from 
#this season. Because they play most tournaments on the same courses every year and most 
#top golfers have little year-to-year variation, I feel comfortable claiming that the 
#top golfers on the PGA Tour are always above average putters for any season from 
#the past and in the future. 
```

\newpage

## References
(1) <https://www.espn.com/golf/stats/player/_/season/2021>
(2) <https://www.pgatour.com/stats.html>
(3) <https://www.pgatour.com/news/2016/05/31/strokes-gained-defined.html>
